A live coding environment for a data exploration language

### 2019-09-23 wip-live-programming/paper-v2.tex
Updating to programming template, tweaks

\title{Live coding environment for a data exploration language}
\author{Tomas Petricek}

@@

A growing amount of code is written to explore and analyze data, but the way data analysts
work with code differs from the way software engineers do. First, many data
scientists write code that introduces few or no abstractions. Second, data exploration is an
interactive process and it blurs the traditional distinction between development-time and run-time.
This poses an interesting challenge for programming language research, because it allows us to
revisit a number of traditional assumptions about programming languages and focus our attention
on a new kind of programming tools that data analysts need.

In this paper, we present \emph{data exploration calculus}, a formally tractable
language that captures the structure of simple data analyses as done, for example, by journalists
exploring open government data. We implement a programming environment for data exploration that
evaluates code instantaneously during editing and shows previews of the results,
while allowing the user to modify code in an unrestricted way in a text
editor. Supporting a text editor is tricky as any edit can change the structure
of code and fully recomputing program after every change would be too expensive.
We present a new technique that allows evaluating and type-checking code while it is
being written while reusing results of previous runs. We formalize the technique using the data
exploration calculus, prove that it is correct and specify when previous results are reused.
Finally, we also illustrate the practicality of our approach using empirical evaluation
and a case study.

As data analysis becomes ever more important use of programming, research on programming languages
and tools needs to consider new kinds of code that people need to write and new kinds of tools
that can support them. The present paper is one step in this important direction.
\end{abstract}
\maketitle

@@

\section{Introduction}
\label{sec:intro}
One of the aspects that make spreadsheets easier to use than programming tools is their
liveness. When you change a value in a cell in Excel, the whole spreadsheet updates instantly
and you immediately see new results, without having to explicitly trigger re-computation
and without having to wait for an extensive period of time.

Increasing number of programming environments aim to provide the same live development experience
for standard programming languages, but doing this is not easy. Fully recomputing the whole program
after every keystroke is inefficient and calculating how a change in the source code changes the
result is extremely hard when the text editor allows arbitrary changes.
Consider the following snippet that gets the release years of 10 most expensive movies from a
data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} =\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
\end{array}
\end{equation*}
%
A live coding environment computes and shows the list of years, but then the programmer modifies the code by
making the constant $\num{10}$ a variable and changing the date format to see the full date:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\ident{count}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking first 10 elements) and only evaluate the
\ident{map} operation to differently format the release dates of already computed top 10 movies.
We present such a live coding environment, describe it formally and empirically evaluate it.
We discuss related work in Section~\ref{sec:future}, but we review the most important directions
next, in order to situate our contributions.

\subsection{Related work}
We focus on simple data exploration as done, for example, by journalists \cite{ddj}
and we aim to bring the experience of using transparent and reproducible programming tools
closer to that of using spreadsheets. When a user edits a cell in a spreadsheet, the results
automatically update. Recomputation in spreadsheets is well-studied,~e.g. by \citet{spreadsheet},
but it is simpler than in our case, because editing is limited to individual cells.

Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some,
can even export the workflow as a script \cite{wrangler}, but data analysts who prefer working with
code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
Those are text-based, but have a limited model of recomputation. Users structure code in
cells and manually reevaluate cells. Many notebook systems do not track dependencies
between cells, which can lead to inconsistencies; an issue that has been addressed
by e.g.~\citet{dataflow,noworkflow} or \citet{wrattler}.

We aim to keep the transparency and reproducibility of notebooks, but add a correct and
efficient instantaneous evaluation mechanism. Existing text-based editors that provide instant,
such as Lighttable \cite{lighttable} and Chrome DevTools, often work only in certain
situations or require full recomputation. A more principled approach has been used by
systems based on structured editing \cite{livenut,lamdu}, which benefit from the fact that code
can only be modified via known operations with known effect on the computation graph
(``extract variable'' has no effect on the result; ``change constant value''
forces recomputation of subsequent code). However, we aim to cater for the many users who
prefer to edit programs as free-form text.

The technical aspects of our technique for providing instantaneous feedback
are related to the work on incremental computation \cite{selfadjusting,incremental}. This focuses
on cases where \emph{data} changes but program stays the same, while our focus is on cases where
\emph{program} changes but data stays. Yet, the technical aspects are related and,
as noted in Section~\ref{sec:future}, language that supports incremental computation could be used
to implement the kind of system presented in this paper.

\subsection{Contributions}
We present the design and implementation of a live coding environment for a data exploration
language that provides a correct and efficient instantaneous feedback, yet, is integrated into an
ordinary text editor. Our most important contributions are:

\begin{itemize}[itemsep=3pt]
\item We introduce \emph{data exploration calculus} (Section~\ref{sec:calculus}), a small formally
  tractable programming language for data exploration. The language is motivated by our review
  of how data analysts work (Section~\ref{sec:background}) and captures the idea that scripts are
  often lists of commands with few or no abstractions with most logic provided by external libraries.

\item Implementing a live coding environment requires a different way of thinking than the one
  presented in classic compiler literature. We capture the essence of the new perspective
  (Section~\ref{sec:formal}) and use it to build our \emph{live preview} mechanism
  (Section~\ref{sec:previews}) evaluates code instantaneously during editing.

@@

\item We extend our core environment in two ways, First, we add instantaneous type checker
  that relies on the same core infrastructure as live previews and enables richer user experience
  through \emph{type providers} (Section~\ref{sec:extra-types}). Second, we add an abstraction mechanism
  that preserves the live coding experience (Section~\ref{sec:extra-abstraction}).
\end{itemize}

@@

\begin{figure}[b]
\includegraphics[scale=0.5]{notebook.png}
\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
  data from the UN trade database (stored in a local file) and adds ISO country codes (from
  an Excel spreadsheet).}
\label{fig:ft-uncomtrade}
\end{figure}

@@

Data scientists often use general-purpose programming languages such as Python, but the kind of
code they write and the way they interact with the system is very different from how software
engineers work \cite{workflow}. This paper focuses on simple data wrangling and data exploration as done, for
example, by journalists analysing government datasets. This section illustrates how such data
analyses look and provides a justification for the design of our data exploration calculus.

@@

Data analysts increasingly use notebook systems such as Jupyter, which
make it possible to combine text, equations and code with results of running the code such as tables
or visualizations. Notebooks blur the distinction between development and execution.
Data analysts write small snippets of code, run them to see results immediately and then revise
them.

Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
models of physical systems to journalists who load data, perform simple aggregations and create visualizations. Our
initial focus is on simple use cases, such as those of journalists. Tooling that makes notebook
systems more interactive and spreadsheet-like can encourage users to choose programming tools
over spreadsheets and produce reproducible and transparent data analyses.

As an example, the analysis done by the Financial Times for an article on plastic waste
\cite{ftnotebooks,ftarticle} joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
an extract from one notebook from the data analysis. The code has a number of remarkable properties:
%
\begin{itemize}[itemsep=3pt]
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options.
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

@@

\item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter
  needs to be set to handle Namibia correctly. These are discovered interactively by writing and
  running a code snippet and examining the output, so providing a rapid feedback is essential.
\end{itemize}

@@

\begin{figure}[b]
  \centering
  \subcaptionbox{
    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. The analyst selects
      Rio 2016 games and counts the number of distinct events per athlete. After typing `.',
      a type provider offers further aggregation operations.
    \label{fig:gamma-dot}
  }{\includegraphics[scale=0.26]{gamma1.png}}
  \subcaptionbox{
    \raggedright\footnotesize Live preview in The Gamma programming environment developed based on
      the theory presented in this paper. The table is produced as
      the data analyst types and updates on-the-fly to show result at the current cursor position.
    \label{fig:gamma-preview}
  }{\includegraphics[scale=0.26]{gamma2.png}}
  \caption{\small Data exploration in The Gamma -- using a type provider (left) and with a live
    preview (right).}
  \label{fig:gamma-intro}
\end{figure}

@@

Data exploration of the kind discussed in the previous section has been the motivation for a
recent work by \citet{gamma} that develops a small scripting language called The Gamma. The language
follows the structure discussed above. A script is a sequence of commands that can either define a
variable or produce an output. The language does not support top-level function declarations
and lambda functions can be used only as method arguments.

Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
such as JavaScript. The Gamma also supports type providers \cite{providers-fsharp},
which can be used to generate types, on the fly, for accessing and exploring external data sources.
Type providers provide object types with members and The Gamma makes using those convenient by
providing auto-complete when the user types the dot symbol (`.') to access a member.

As illustrated by \citet{gamma}, the combination of type providers and auto-complete on~`.' makes it
possible to complete a large number of data exploration tasks through a very simple interaction of
selecting operations from a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on
Olympic medals using a type provider. The names of identifiers such as \qident{sum Bronze} do not have
any special meaning. They are merely members generated by the type provider, based on information
about the data source. The type provider used in this example generates an object with members for
different data transformations, such as \qident{group data}, which return further objects with
members for specifying transformation parameters, such as selecting the grouping key using
\qident{by Athlete}.

The Gamma language is richer, but the example in Figure~\ref{fig:gamma-dot} shows that non-trivial
data exploration can be done using a very simple language. The assumptions about structure of
code that are explicit in The Gamma are implicitly present in Python and R data analyses
produced by journalists, economists and other users with other than programming background.
When we refer to The Gamma in this paper, readers not familiar with it can assume that we focus
on a subset of Python.

@@

The practical implementation work that complements the theory presented in this paper implements a live
coding environment for The Gamma and is discussed in Section~\ref{sec:evaluation-case} in detail. Earlier work by
\citet{gamma} discussed The Gamma language and the type provider mechanism, but the implementation
provided only a standard text editor with auto-completion. Our work brings the experience of using
The Gamma closer to that of spreadsheets by developing an editor that evaluates scripts
instantaneously in background and displays the results as \emph{live previews}.

An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
Gamma program consist of a list of commands which are either expressions or let bindings. Our
editor displays a live preview below the command that the user is currently editing. The preview
shows the result of evaluating the expression or the value assigned to a bound variable. When the
user changes the code, the preview is updated automatically, without any additional interaction
with the user.

There is a number of guiding principles that our work follows. First, we allow the analyst
to edit code in an unrestricted form in a text editor. Although structured editors provide an
appealing alternative and make recomputation easier, we prefer the transparency
and openness of plain text. Second, we focus on the scenario when code changes, but input
does not. Rapid feedback allows the analyst to quickly adapt code to correctly handle
corner cases that typical analysis involves. In contrast to work on incremental computation,
we do not consider the case when data changes, although supporting interactive data exploration
of streaming data is an interesting future direction.

@@

The \emph{data exploration calculus} is a small, formally tractable language for data exploration.
The calculus is not Turing-complete and it can only be used together with external libraries that
define what objects are available and what the behaviour of their members is. This is
sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
define the calculus in this section and then use it as the basis for discussing our live preview
mechanism in Section~\ref{sec:formal}.

We initially present the data exploration calculus without a static type system as the live preview
mechanism does not rely on types. We consider type checking later in Section~\ref{sec:extra-types}
as the mechanism that is used for live previews can also be used for instantaneous type checking.

@@

\begin{figure}
\raggedright
\hspace{1.5em}{\small\sffamily Programs, commands, terms, expressions and values}
\begin{equation*}
\begin{array}{lcl}
p &\narrow{=}& c_1; \ldots; c_n\\
c &\narrow{=}& \kvd{let}~x = t \lsep t\\
\end{array}
\qquad
\begin{array}{lcl}
t &\narrow{=}& o \lsep x \\
  &\narrow{|}& t.m(e, \ldots, e)
\end{array}
\qquad
\begin{array}{lcl}
e &\narrow{=}& t \lsep \lambda x\rightarrow e\\
v &\narrow{=}& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

@@

The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ (one object for each command, including let bindings) which can be
displayed as intermediate results of the data analysis. The operational semantics is parameterized
by a relation $\rightsquigarrow_\epsilon$ that models the functionality of the external libraries
used with the calculus and defines reduction rules for member accesses. The relation has the
following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
The operation is invoked on an object and takes values (objects or function values) as arguments.
It always results in an opaque object.
Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evaluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
The rule \rname{external} applies reductions provided by an external library in a
call-by-value order and \rname{let} substitutes a value of an evaluated variable in
all subsequent commands. Note that \rname{let} leaves the value of the variable
in the resulting list of commands.

\subsection{Properties}
The data exploration calculus is a very simple language and it has a number of desirable
properties. However, some of those require that the relation $\rightsquigarrow_\epsilon$, which
defines evaluation for external libraries, satisfies a number of conditions. Our main motivation
is to capture properties that allow us to prove that our method of evaluating live previews,
discussed in Section~\ref{sec:evaluation-correctness}, is correct.

@@

\noindent
We say that two expressions
$e$ and $e'$ are observationally equivalent if, for any context $C$, the expressions $C[e]$ and
$C[e']$ reduce to the same value. Lambda functions $\lambda x\!\rightarrow\!2$ and
$\lambda x\!\rightarrow\!1\!+\!1$ are not equal, but they are observationally equivalent.
We require that external libraries satisfy two conditions. First, when a method is called with
observationally equivalent values as arguments, it should return the same value. Second, the
evaluation of $o.m(v_1,\ldots,v_n)$ should be defined for all $o, i$ and $v_i$. The external
library will typically satisfy this by defining an error object and reducing all invalid calls
to the error object, following the standard method of \citet{gowrong}.

\begin{definition}[External library]
\label{def:external}
An external library consists of a set of objects $O$ and a reduction relation
$\rightsquigarrow_\epsilon$ that satisfies the following two properties:
\begin{itemize}
\item[-] \emph{Completeness.} For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
  that $o.m(v_1, \ldots, v_i) \rightsquigarrow_\epsilon o'$.
\vspace{0.5em}
\item[-] \emph{Compositionality.} For observationally equivalent arguments, the reduction returns
  the same object, i.e.~given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
  $e_0.m(e_1, \ldots, e_n) \rightsquigarrow^{*} o$ and $e'_0.m(e'_1, \ldots, e'_n) \rightsquigarrow^{*} o'$ then
  if for any contexts $C_0, C_1, \ldots, C_n$ it holds that if $C_i[e_i] \rightsquigarrow^{*} o_i$ and
  $C_i[e'_i] \rightsquigarrow^{*} o_i$ for some $o_i$ then $o = o'$.
\end{itemize}
\end{definition}

\noindent
The compositionality condition is essential for proving the correctness of our live preview mechanism.
Completeness allows us to prove normalization, i.e.~all programs reduce to
a value -- although the resulting value may be an error value provided by the external library.

@@

\section{Formalising live coding environment}
\label{sec:formal}

A naive way of providing live previews during code editing would be to re-evaluate the code
after each change. This would be wasteful -- for example, when writing
code, most changes are additive and the preview can be updated by evaluating just the newly added
code. In this section, we develop foundations for an efficient way of displaying live previews
for the data exploration calculus.

@@

The key idea behind our method is to maintain a dependency graph \cite{dependencies} with
nodes representing individual operations of the computation that can be (partially) evaluated
to obtain a preview. Each time the program text is modified, we parse it afresh (using an
error-recovering parser) and bind the abstract syntax tree to the dependency graph.
When binding a new expression to the graph, we reuse previously created nodes as long as
they have the same dependencies. For expressions that have a new structure, we create new nodes.

>>>>

\title{A live coding environment for a data exploration language}
\author{Tomas Petricek}

@@

A growing amount of code is written to explore and analyze data, but the way data analysts
work with code differs from the way software engineers do. First, many data
scientists write code that introduces few or no abstractions. Second, data exploration is an
interactive process that blurs the traditional distinction between development-time and run-time.
This poses an interesting challenge for programming language research, inviting us to
revisit a number of traditional assumptions about programming languages and design a new kind of
programming tools that are suited for data analyst needs.

We present the \emph{data exploration calculus}, a formal language that captures the structure of
scripts written to explore data. We focus on the simplest kind of data analyses such as those
written by journalists to analyse open government data. We implement a programming environment,
based on this calculus, that evaluates code instantly during editing and shows previews of the results,
while allowing the user to modify code in an unrestricted way in a text editor. Supporting interactive
editing is tricky. Any edit can change the structure of code and fully recomputing the output would
be too expensive. We present a technique that allows evaluating and type-checking code while it is
being written and that reuses results of previous runs. We formalize the technique using the data
exploration calculus, prove that it is correct and specify when previous results are reused.
Finally, we illustrate the practicality of our approach using empirical evaluation
and a case study.

As data analysis becomes an ever more important use of programming, research on programming languages
and tools needs to consider new kinds of programming workflows appropriate for those domains and
conceive new kinds of tools that can support them. The present paper is one step in this important direction.
\end{abstract}
\maketitle

@@

\section{Introduction}
\label{sec:intro}
One of the aspects that make spreadsheets easier to use than other programming tools is their
liveness. When you change a value in a cell in Excel, the whole spreadsheet updates instantly
and you immediately see new results, without having to explicitly trigger re-computation
and without having to wait for an extensive period of time.

Increasing numbers of programming environments aim to provide a live development experience
for standard programming languages, but doing this is not easy. Fully recomputing the whole program
after every keystroke is inefficient. Calculating how a change in the source code changes the
result is extremely hard when the text editor allows arbitrary changes.
Consider the following snippet that gets the release years of the 10 most expensive movies from a
data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} =\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
\end{array}
\end{equation*}
%
A live coding environment computes and shows the list of years. Suppose that the programmer then modifies the code by
making $\num{10}$ a variable and changing the date format to see the full date:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\ident{count}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking first 10 elements) and only evaluate the
\ident{map} operation to differently format the release dates of already computed top 10 movies.
This can already be done in a number of systems based on structured code editors, but our
approach allows this behaviour using a standard text editor. We discuss related work in more detail in
Section~\ref{sec:future}, but we review the most important directions here, in order to
situate our contributions.

\subsection{Related work}
Simple data exploration performed, for example, by journalists \cite{ddj} is done either
programmatically or using spreadsheets. The former is reproducible and transparent, but requires
expert programming skills. The latter is easy, but it limits reproducibility and is error-prone.
We aim to bring some aspects of spreadsheets, especially liveness, to programmatic data exploration.
Recomputation in spreadsheets is well studied,~e.g. by \citet{spreadsheet}, but extending it to
code written and edited in an ordinary text editor poses interesting challenges.

Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some
can even export the workflow as a script \cite{wrangler}. Data analysts who prefer working with
code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
Those are text-based, but have a limited model of recomputation. Users structure code in
cells and manually reevaluate cells. Many notebook systems do not track dependencies
between cells, which can lead to inconsistencies, an issue that has been addressed
by e.g.~\citet{dataflow,noworkflow} and \citet{wrattler}.

Existing text-based editors that provide instant feedback,
such as Lighttable \cite{lighttable} and Chrome DevTools, often work only in certain
situations or require full recomputation. A more principled approach can be used by
systems based on structured editing \cite{livenut,lamdu}, which benefit from the fact that code
can is modified via known operations with known effect on the computation graph
(e.g.~``extract variable'' has no effect on the result; ``change constant value''
forces recomputation of subsequent code). However, we aim to cater for the many users who
prefer to edit programs as free-form text.
%
The technical aspects of our technique for providing instant feedback
are related to the work on incremental computation \cite{selfadjusting,incremental}. This focuses
on cases where \emph{data changes} but program stays the same, while our focus is on cases where
\emph{program changes} but data stays. Yet, the technical aspects are related and,
as noted in Section~\ref{sec:future}, a language that supports incremental computation could be used
to implement the kind of system presented in this paper.

\subsection{Contributions}
We present the design and implementation of a live coding environment for a data exploration
language that provides correct and efficient instant feedback, yet is integrated into an
ordinary text editor. Our most important contributions are:

\begin{itemize}[itemsep=3pt]
\item We introduce the \emph{data exploration calculus} (Section~\ref{sec:calculus}), a small formally
  tractable language for data exploration. The calculus is motivated by our review
  of how data analysts work (Section~\ref{sec:background}) and makes it possible to write simple,
  transparent and reproducible data analyses.

\item Implementing a live coding environment requires a different way of thinking than the one
  presented in classic compiler literature, because it does not operate in a batch mode. We capture
  the essence of the new perspective (Section~\ref{sec:formal}) and use it to build our \emph{live
  preview} mechanism (Section~\ref{sec:previews}) that evaluates code instantly during editing.

@@

\item We describe two extensions of our core environment. First, we add instant type checker
  that relies on the same core infrastructure as live previews and enables richer user experience
  through \emph{type providers} (Section~\ref{sec:extra-types}). Second, we add a code reuse mechanism
  that supports the live coding experience (Section~\ref{sec:extra-abstraction}).
\end{itemize}

@@

\begin{figure}[b]
\vspace{-0.5em}
\includegraphics[scale=0.5]{notebook.png}
\vspace{-0.5em}
\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
  data from the UN trade database (stored in a local file) and adds ISO country codes (from
  an Excel spreadsheet).}
\label{fig:ft-uncomtrade}
\vspace{-0.5em}
\end{figure}

@@

Data scientists often use general-purpose programming languages such as Python, but the kind of
code they write and the way they interact with the system is very different from how software
engineers work. The different nature of data science workflow has been documented
by \citet{workflow}. This paper focuses on simple data wrangling and data exploration as done, for
example, by journalists analysing government datasets. This section illustrates how such data
analyses typically look and provides a justification for the design of our data exploration calculus.

@@

Data analysts increasingly use notebook systems such as Jupyter, which
make it possible to combine text, equations and code with results of running the code, such as tables
or visualizations. Notebooks blur the distinction between development and execution.
Data analysts write small snippets of code, run them to see results immediately and then revise
them.

Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
models of physical systems to journalists who load data, perform simple aggregations and create visualizations. Our
initial focus is on the simplest use cases. We believe that making programmatic
data exploration more interactive and spreadsheet-like can encourage users to choose programming tools
over spreadsheets, allowing them to produce reproducible and transparent data analyses.

As an example, the analysis done by the Financial Times for an article on plastic waste
\cite{ftnotebooks,ftarticle} joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
an extract from one notebook of the data analysis. The code has a number of important properties:
%
\begin{itemize}[itemsep=3pt]
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options.
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

@@

\item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter
  needs to be set to handle Namibia correctly. These are discovered interactively by writing and
  running a code snippet and examining the output, so providing an instant feedback is essential.
\end{itemize}

@@

\begin{figure}[b]
  \centering
  \subcaptionbox{
    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. The analyst selects
      Rio 2016 games and counts the number of distinct events per athlete. After typing `.'
      a type provider offers further aggregation operations.
    \label{fig:gamma-dot}
  }{\includegraphics[scale=0.26]{gamma1.png}}
  \subcaptionbox{
    \raggedright\footnotesize
      Our live coding environment for The Gamma language based on the theory presented in this paper.
      The table is produced as
      the data analyst types and updates on-the-fly to show result at the current cursor position.
    \label{fig:gamma-preview}
  }{\includegraphics[scale=0.26]{gamma2.png}}
  \caption{\small Previous work on The Gamma by \citet{gamma} (left) and our new editor with live preview (right).}
  \label{fig:gamma-intro}
\end{figure}

@@

Data exploration of the kind discussed in the previous section has been the motivation for
recent work by \citet{gamma} that develops a small scripting language called The Gamma.
Scripts in The Gamma are sequences of commands that can either define a variable or produce an
output. The language does not support top-level function declarations
and lambda functions can be used only as method arguments.

Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
such as JavaScript. The Gamma supports type providers \cite{providers-fsharp},
which generate types on the fly, for accessing and exploring external data sources.
Type providers provide object types with members and The Gamma makes using those convenient by
providing auto-complete when the user types the dot symbol (`.') to access a member.

As illustrated by \citet{gamma}, the combination of type providers and auto-complete on~`.' makes it
possible to complete a large number of data exploration tasks through the very simple interaction of
selecting operations from a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on
Olympic medals using a type provider. Identifiers such as \qident{sum Bronze}
are merely names of members generated by the type provider, based on information
about the data source. The type provider used in this example generates an object with members for
different data transformations, such as \qident{group data}, which return further objects with
members for specifying transformation parameters, such as selecting the grouping key using
\qident{by Athlete}.

The Gamma language is richer, but the example in Figure~\ref{fig:gamma-dot} shows that non-trivial
data exploration can be done using a very simple language. The assumptions about structure of
code that are explicit in The Gamma are implicitly present in Python and R data analyses
produced by journalists, economists and other users with other than programming background.
When we refer to The Gamma in this paper, readers not familiar with it can consider a subset of Python.

@@

The implementation that accompanies this paper builds a live coding environment for The Gamma
language. The implementation is discussed in Section~\ref{sec:evaluation-case} and it replaces
the text editor with auto-complete, originally used by The Gamma, with a live coding environment
that provides \emph{live previews}. We choose to extend The Gamma over using a subset of Python
for its simplicity as it lets us focus on the principles of live coding systems.

An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
Gamma programs consist of lists of commands which are either expressions or let bindings. Our
editor displays a live preview below the command that the user is currently editing. The preview
shows the result of evaluating the expression or the value assigned to a bound variable. When the
user changes the code, the preview is updated automatically, without any additional interaction
with the user.

There are a number of guiding principles that inform our design. First, we allow the analyst
to edit code in an unrestricted form in a text editor. Although structured editors provide an
appealing alternative and make recomputation easier, we prefer the flexibility of plain text.
Second, we focus on the scenario when code changes, but input
does not. Rapid feedback allows the analyst to quickly adapt code to correctly handle
corner cases that typical analysis involves. In contrast to work on incremental computation,
we do not consider the case when data changes, although supporting interactive data exploration
of streaming data is an interesting future direction.

@@

The \emph{data exploration calculus} is a small formal language for data exploration.
The calculus is not, in itself, Turing-complete and it can only be used together with external libraries that
define what objects are available and what the behaviour of their members is. This is
sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
define the calculus in this section and then use it as the basis for discussing our live preview
mechanism in Section~\ref{sec:formal}.

We initially present the data exploration calculus without a static type system as the live preview
mechanism does not rely on types. We consider type checking later in Section~\ref{sec:extra-types}.
Interestingly, the mechanism that is used for live previews can also be used for instant type checking.

@@

\begin{figure}
\raggedright
\hspace{1.5em}{\small\sffamily Programs, commands, terms, expressions and values}
\begin{equation*}
\begin{array}{lcl}
p &\narrow{::=}& c_1; \ldots; c_n\\
c &\narrow{::=}& \kvd{let}~x = t \lsep t\\
\end{array}
\qquad
\begin{array}{lcl}
t &\narrow{::=}& o \lsep x \\
  &\narrow{|}& t.m(e, \ldots, e)
\end{array}
\qquad
\begin{array}{lcl}
e &\narrow{::=}& t \lsep \lambda x\rightarrow e\\
v &\narrow{::=}& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

@@

The data exploration calculus is a call-by-value language. We model evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ (one object for each command, including let bindings) which can be
displayed as intermediate results of the data analysis. The operational semantics is parameterized
by a relation $\rightsquigarrow_\epsilon$ that models the functionality of the external libraries
used with the calculus and defines the reduction behaviour for member accesses. The relation has the
following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
Here, the operation $m$ is invoked on an object and takes values (objects or function values) as arguments.
The reduction always results in an object.
Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evaluation of a command and $C_p$ defines left-to-right evaluation of a program.
The rule \rname{external} calls a method provided by an external library in a
call-by-value fashion and \rname{let} substitutes a value of an evaluated variable in
all subsequent commands and leaves the result in the list of commands.
Note that our semantics does not define how $\lambda$ functions are reduced. This is done by
external libraries, which will typically supply functions with arguments using standard $\beta$-reduction.
The behaviour is subject to constraints discussed next.

\subsection{Properties}
The data exploration calculus has a number of desirable properties. Some of those require that the
relation $\rightsquigarrow_\epsilon$, which defines evaluation for external libraries, satisfies a
number of conditions. We discuss \emph{normalization} and \emph{let elimination} in this section.
Those two are particularly important as they will allow us to prove correctness of our method of
evaluating live previews in Section~\ref{sec:evaluation-correctness}.

@@

\noindent
We say that two expressions
$e$ and $e'$ are \emph{observationally equivalent} if, for any context $C$, the expressions $C[e]$ and
$C[e']$ reduce to the same value. Lambda functions $\lambda x\!\rightarrow\!2$ and
$\lambda x\!\rightarrow\!1\!+\!1$ are not equal, but they are observationally equivalent.
We require that external libraries satisfy two conditions. First, when a method is called with
observationally equivalent values as arguments, it should return the same value. Second, the
evaluation of $o.m(v_1,\ldots,v_n)$ should be defined for all $o, i$ and $v_i$. The external
library will typically satisfy this by defining an error object and reducing all invalid calls
to the error object, following the standard method of \citet{gowrong}.

\begin{definition}[External library]
\label{def:external}
An external library consists of a set of objects $O$ and a reduction relation
$\rightsquigarrow_\epsilon$ that satisfies the following two properties:
\begin{itemize}
\item[-] \emph{Totality.} For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
  that $o.m(v_1, \ldots, v_i) \rightsquigarrow_\epsilon o'$.
\vspace{0.5em}
\item[-] \emph{Compositionality.} For observationally equivalent arguments, the reduction returns
  the same object, i.e.~given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
  $e_0.m(e_1, \ldots, e_n) \rightsquigarrow^{*} o$ and $e'_0.m(e'_1, \ldots, e'_n) \rightsquigarrow^{*} o'$ then
  if for any contexts $C_0, C_1, \ldots, C_n$ it holds that if $C_i[e_i] \rightsquigarrow^{*} o_i$ and
  $C_i[e'_i] \rightsquigarrow^{*} o_i$ for some $o_i$ then $o = o'$.
\end{itemize}
\end{definition}

\noindent
The compositionality condition is essential for proving the correctness of our live preview mechanism.
Totality allows us to prove normalization, i.e.~all programs reduce to
a value -- although the resulting value may be an error value provided by the external library.

@@

\section{Formalising a live coding environment}
\label{sec:formal}

A naive way of providing live previews during code editing would be to re-evaluate the code
after each change. This would be wasteful -- when writing code to explore data, most changes are
additive and the preview can be updated by evaluating just the newly added
code. In this section, we develop foundations for an efficient way of displaying live previews
for the data exploration calculus.

@@

The key idea behind our method is to maintain a dependency graph \cite{dependencies} with
nodes representing individual operations of the computation that can be evaluated
to obtain a preview. Each time the program text is modified, we parse it afresh (using an
error-recovering parser) and bind the abstract syntax tree to the dependency graph.
When binding a new expression to the graph, we reuse previously created nodes as long as
they have the same dependencies. For expressions that have a new structure, we create new nodes.
